Entropy-SGD: Biasing Gradient Descent Into Wide Valleys

### 2017-01-11 doc/chaudhari.choromanska.ea.iclr17.tex
wip on sec. 4.3

Another related area is homotopy continuation methods that convolve the loss function to solve sequentially refined optimization problems~\citep{allgower2012numerical,mobahi2015link} or methods that perturb the weights or activations to average the gradient~\citep{gulcehre2016mollifying} with an aim to smooth the rugged energy landscape. Such a smoothing is however different from local entropy. For instance, the latter places more weight on wide local minima even if they are much shallower than the global minimum (cf. Fig.~\ref{fig:entropyfig}) which cannot be obtained by smoothing. In fact, smoothing can introduce an artifical minimum between two sharp valleys connected by a high energy barrier which is detrimental to generalization. Please refer to Sec.~\ref{ss:connection_variational} for a more elaborate discussion. In order to be effective, continuation techniques also require that minimizers of successively smaller convolutions of the loss function lie close to each other~\citep{DBLP:conf/icml/HazanLS16}; it is not clear whether this is true for deep networks. Local entropy exploits wide minima which have been shown to exist in a variety of learning problems~\citep{monasson1995weight,cocco1996analytical}; our experiments on the eigenspectrum of the Hessian are a step to showing the existence of such wide minima in modern deep networks.

@@

The scoping parameter $\g$ is fixed in Alg.~\ref{alg:entropysgd}. For large values of $\g$, the SGLD updates happen in a small neighborhood of the current parameters $x$ while low values of $\g$ allow the inner SGLD to explore further away from $x$. A ``reverse-annealing'' of the scope $\g$, i.e. increasing $\g$ as training progresses, thus has the effect of exploring the energy landscape on progressively finer scales. We call this process ``scoping'' and use a simple exponential schedule given by
$$
    \g(t) = \g_0\ \rbrac{1 + \g_1}^t;
$$
for the $t^{\trm{th}}$ parameter update. We have experimented with linear, quadratic and bounded exponential ($\g_0 \rbrac{1 - e^{-\tau t}}$) scoping schedules with similar qualitative results.

@@

\pc{We can show that $\entropysgd$ results in a smoother loss function and obtains better generalization error than the original objective~\eqref{eq:dnn_objective} if trained for the same number of iterations. Proofs are deferred to the Appendix (see Sec.~\ref{s:app:proofs}).}{clarify that the analysis here is only a sketch, it is not rigorous}

With some overload of notation, we assume that the original loss $f(x)$ is $\b$-smooth, i.e., for all $x,y \in \reals^n$, we have $\norm{\nabla f(x) - \nabla f(y)} \leq \b\ \norm{x-y}$. We additionally assume for the purpose of analysis that no eigenvalue of the Hessian $\nabla^2 f(x)$ lies in the set $[-2 \g-c, c]$ for some small $c > 0$.
%
\begin{lemma}
\label{lem:smoothness_reduction}
The objective $F(x, \g;\ \Xi)$ in~\eqref{eq:entropysgd_objective} is $\f{\a}{1 + \g^{-1}\ c}$-Lipschitz and $\f{\b}{1 + \g^{-1}\ c}$-smooth.
\end{lemma}
The local entropy objective is thus smoother than the original objective.

Let us now obtain a bound on the improvement in generalization error. We denote an optimization algorithm, viz., SGD or $\entropysgd$ by $A(\Xi)$, it is a function of the dataset $\Xi$ and outputs the parameters $x^*$ upon termination. Stability of the algorithm~\citep{bousquet2002stability} is then a notion of how much its output differs in loss upon being presented with two datasets $\Xi$ and $\Xi'$ that differ in at most one sample,
$$
    \sup_{\xi\ \in\ \Xi\ \cup\ \Xi'}\ \sqbrac{ f \rbrac{A(\Xi), \xi} - f \rbrac{A(\Xi'), \xi}} \leq \e.
$$
\citet{hardt2015train} connect uniform stability to generalization error and show that an $\e$-stable algorithm $A(\Xi)$ has generalization error bounded by $\e$, i.e., if $A(\Xi)$ terminates with parameters $x^*$,
$$
    \abs{ \E_{\Xi} \rbrac{R_\Xi(x^*) - R(x^*)} } \leq \e;
$$
where the left hand side is the generalization error: it is the difference between the empirical loss $R_\Xi(x) := \f{1}{N}\ \sum_{k=1}^N\ f(x, \xi_k)$ and the population loss $R(x) := \E_{\xi}\ f(x, \xi)$.

@@

\subsection{Connection to variational inference}
\label{ss:connection_variational}
The local entropy based objective in $\entropysgd$ has similar motivations as variational inference.

@@

\paragraph{Comparison with SGLD:}
\todo{A common way to train Bayesian neural networks is through MCMC algorithms to approximate the likelihood of data marginalized over the weights. Indeed, SGLD which forms the inner loop of $\entropysgd$ is a popular algorithm for variational inference~\citep{balan2015bayesian}. We run vanilla SGLD for two prototypical networks in Sec.~\ref{s:expt}, $\lenet$ on MNIST and $\allcnn$ on CIFAR-10.}

@@

>>>>

Another related area is homotopy continuation methods that convolve the loss function to solve sequentially refined optimization problems~\citep{allgower2012numerical,mobahi2015link} or methods that perturb the weights or activations to average the gradient~\citep{gulcehre2016mollifying} with an aim to smooth the rugged energy landscape. Such a smoothing is however different from local entropy. For instance, the latter places more weight on wide local minima even if they are much shallower than the global minimum (cf. Fig.~\ref{fig:entropyfig}) which cannot be obtained by smoothing. In fact, smoothing can introduce an artifical minimum between two sharp valleys connected by a high energy barrier which is detrimental to generalization. Please refer to Sec.~\ref{ss:app:connection_variational} for a more elaborate discussion. In order to be effective, continuation techniques also require that minimizers of successively smaller convolutions of the loss function lie close to each other~\citep{DBLP:conf/icml/HazanLS16}; it is not clear whether this is true for deep networks. Local entropy exploits wide minima which have been shown to exist in a variety of learning problems~\citep{monasson1995weight,cocco1996analytical}; our experiments on the eigenspectrum of the Hessian are a step to showing the existence of such wide minima in modern deep networks.

@@

The scoping parameter $\g$ is fixed in Alg.~\ref{alg:entropysgd}. For large values of $\g$, the SGLD updates happen in a small neighborhood of the current parameters $x$ while low values of $\g$ allow the inner SGLD to explore further away from $x$. A ``reverse-annealing'' of the scope $\g$, i.e. increasing $\g$ as training progresses, thus has the effect of exploring the energy landscape on progressively finer scales. We call this process ``scoping'' and use a simple exponential schedule given by
$$
    \g(t) = \g_0\ \rbrac{1 + \g_1}^t;
$$
for the $t^{\trm{th}}$ parameter update. We have experimented with linear, quadratic and bounded exponential ($\g_0 \rbrac{1 - e^{-\tau t}}$) scoping schedules and obtained qualitatively similar results.

@@

We can show that $\entropysgd$ results in a smoother loss function and obtains better generalization error than the original objective~\eqref{eq:dnn_objective} if trained for the same number of iterations. Proofs are deferred to the Appendix (see Sec.~\ref{s:app:proofs}).

With some overload of notation, we assume that the original loss $f(x)$ is $\b$-smooth, i.e., for all $x,y \in \reals^n$, we have $\norm{\nabla f(x) - \nabla f(y)} \leq \b\ \norm{x-y}$. We additionally assume for the purpose of analysis that no eigenvalue of the Hessian $\nabla^2 f(x)$ lies in the set $[-2 \g-c, c]$ for some small $c > 0$.
%
\begin{lemma}
\label{lem:smoothness_reduction}
The objective $F(x, \g;\ \Xi)$ in~\eqref{eq:entropysgd_objective} is $\f{\a}{1 + \g^{-1}\ c}$-Lipschitz and $\f{\b}{1 + \g^{-1}\ c}$-smooth.
\end{lemma}
The local entropy objective is thus smoother than the original objective.
%
Let us now obtain a bound on the improvement in generalization error. We denote an optimization algorithm, viz., SGD or $\entropysgd$ by $A(\Xi)$, it is a function of the dataset $\Xi$ and outputs the parameters $x^*$ upon termination. Stability of the algorithm~\citep{bousquet2002stability} is then a notion of how much its output differs in loss upon being presented with two datasets $\Xi$ and $\Xi'$ that differ in at most one sample,
$$
    \sup_{\xi\ \in\ \Xi\ \cup\ \Xi'}\ \sqbrac{ f \rbrac{A(\Xi), \xi} - f \rbrac{A(\Xi'), \xi}} \leq \e.
$$
\citet{hardt2015train} connect uniform stability to generalization error and show that an $\e$-stable algorithm $A(\Xi)$ has generalization error bounded by $\e$, i.e., if $A(\Xi)$ terminates with parameters $x^*$,
$$
    \abs{ \E_{\Xi} \rbrac{R_\Xi(x^*) - R(x^*)} } \leq \e;
$$
where the left hand side is the generalization error: it is the difference between the empirical loss $R_\Xi(x) := \f{1}{N}\ \sum_{k=1}^N\ f(x, \xi_k)$ and the population loss $R(x) := \E_{\xi}\ f(x, \xi)$.

@@

@@

@@

\todo{
\subsection{Connection to variational inference}
\label{ss:app:connection_variational}
The local entropy based objective in $\entropysgd$ has similar motivations as variational inference.

\paragraph{Comparison with SGLD:}
A common way to train Bayesian neural networks is through MCMC algorithms to approximate the likelihood of data marginalized over the weights. Indeed, SGLD which forms the inner loop of $\entropysgd$ is a popular algorithm for variational inference~\citep{balan2015bayesian}. We run vanilla SGLD for two prototypical networks in Sec.~\ref{s:expt}, $\lenet$ on MNIST and $\allcnn$ on CIFAR-10.
}
